---
title: "MA388 Sabermetrics: Lesson 26"
subtitle: "Statast Spray Charts"
format:
  html:
    theme: cosmo
    toc: true
  pdf:
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
---


```{r}
library(tidyverse)
library(knitr)
library(baseballr)
```


Review

**Last class we talked about the streakiness or clumpiness coefficient.  What does this metric measure and how do we calculate it?**

\vspace{1.in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

**True or False.  If a player had a high batting average for the season, this means that they were not a streaky player.**

\vspace{1.in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

**Could a player be temporally streaky but spatially predictable?**

\vspace{1.in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

What do Joey Gallo, Albert Pujols, and DJ LeMahieu have in common?

*Here's one answer:* They had some of the most extreme spray charts among MLB hitters in 2017.

Of course, all three hitters went to different extremes -- Gallo was one of MLB's most pull-heavy hitters, Pujols hit balls to center field more than almost anyone, and LeMahieu used the opposite field like nobody else in baseball. 

Here is the full article: [Extreme pull and opposite-field hitters (MLB.com, 2017)](https://www.mlb.com/news/statcast-extreme-pull-oppo-hitters-2017-c262687474)

Let's explore their spray charts for that year.  

```{r}
library(tidyverse)
library(baseballr)
``` 

```{r, cache=TRUE}
gallo <- scrape_statcast_savant(start_date = "2017-05-01", 
                                 end_date = "2017-12-31", 
                                 playerid = 608336,
                                 player_type = "batter")

pujols <- scrape_statcast_savant(start_date = "2017-05-01", 
                                 end_date = "2017-12-31", 
                                 playerid = 405395,
                                 player_type = "batter")

lemahieu <- scrape_statcast_savant(start_date = "2017-05-01", 
                                 end_date = "2017-12-31", 
                                 playerid = 518934,
                                 player_type = "batter")

all_data = rbind(gallo, lemahieu, pujols)
```

Filter the data to only include balls in play.

```{r}
all_bip <- all_data |>
  filter(type == "X")

# hc_x and hc_y are the hit coordinates for balls in play.
all_bip |>
  select(player_name, events, hc_x, hc_y) |>
  head()
```

Now let's create the base plot for our spray chart.  

```{r}
spray_chart <- function(...) {
  ggplot(...) + 
    geom_curve(x = 33, xend = 223, y = -100, yend = -100,
               curvature = -.65) +
    geom_segment(x = 128, xend = 33, y = -208, yend = -100) +
    geom_segment(x = 128, xend = 223, y = -208, yend = -100) +
    geom_curve(x = 83, xend = 173, y = -155, yend = -156,
               curvature = -.65, linetype = "dotted") +
    coord_fixed() + 
    scale_x_continuous(NULL, limits = c(25, 225)) + 
    scale_y_continuous(NULL, limits = c(-225, -25))
}
```

**Where would we expect these different events to occur?**  

```{r}
spray_chart(all_bip, aes(x = hc_x, y = -hc_y, color = events)) + 
  scale_color_hue()+
  theme_bw()
```

Let's look at all three players together.  

```{r}
spray_chart(all_bip, aes(x = hc_x, y = -hc_y, color = events)) + 
  geom_point() +
  scale_color_hue()+
  theme_bw()
```

Now let's take a look at each player individually.

```{r}
all_bip |> 
  filter(player_name=="Gallo, Joey") |> 
  spray_chart(aes(x = hc_x, y = -hc_y, color = events)) + 
  geom_point() + 
  scale_color_hue()+
  labs(title = "Spray Chart Joey Gallo (Left Handed)")+
  theme_bw()
```

```{r}
all_bip |> 
  filter(player_name=="Pujols, Albert") |> 
  spray_chart(aes(x = hc_x, y = -hc_y, color = events)) + 
  geom_point() + 
  scale_color_hue()+
  labs(title = "Spray Chart Albert Pujols (Right Handed)")+
  theme_bw()
```

```{r}
all_bip |> 
  filter(player_name=="LeMahieu, DJ") |> 
  spray_chart(aes(x = hc_x, y = -hc_y, color = events)) + 
  geom_point() + 
  scale_color_hue()+
  labs(title = "Spray Chart DJ LeMahieu (Right Handed)")+
  theme_bw()
```

For a side-by-side comparison:

```{r}
all_bip |>
  spray_chart(aes(x = hc_x, y = -hc_y, color = events)) + 
  geom_point() + 
  scale_color_hue()+
  theme_bw()+
  theme(legend.position = "none")+
  facet_wrap(~player_name)
```

```{r}
# Traditional defensive alignment
traditional_defense <- tibble(
  position = c("P", "C", "1B", "2B", "SS", "3B", "LF", "CF", "RF"),
  x = c(128, 128, 160, 140, 115, 95, 80, 128, 180),
  y = c(-165, -220, -150, -135, -135, -150, -80, -60, -80)
)

# Shift vs pull-heavy LEFTY (e.g., Gallo)
shift_defense <- tibble(
  position = c("P", "C", "1B", "2B_shift", "SS_shift", "3B_shift", "LF", "CF", "RF"),
  x = c(128, 128, 165, 155, 135, 110, 90, 135, 185),
  y = c(-165, -220, -150, -115, -125, -135, -85, -65, -85)
)

add_defense <- function(defense_data, color = "black") {
  list(
    geom_point(data = defense_data,
               aes(x = x, y = y),
               color = color,
               size = 3),
    
    geom_text(data = defense_data,
              aes(x = x, y = y, label = position),
              vjust = -1,
              color = color,
              size = 3)
  )
}

all_bip |>
  # filter(player_name == "Gallo, Joey") |> 
  spray_chart(aes(x = hc_x, y = -hc_y, color = events)) + 
  # geom_point(alpha = 0.6) +
  add_defense(traditional_defense, color = "blue") +
  add_defense(shift_defense, color = "red") +
  # labs(title = "Joey Gallo vs Traditional Defense") +
  theme_bw()

```


In general, players hit ground balls to the pull side of the field.  Let's take a look at how these three players hit their ground balls and the percentage of times they faced infield shifts.

```{r}
left_handed_batters = gallo |> 
  filter(launch_angle < 10) |> 
  summarize(N = n(),
            Pulled = 100 * mean(hc_x > 125.42, na.rm = TRUE),
            Shift = 100 * 
              mean(if_fielding_alignment == "Infield shift",
                   na.rm = TRUE),.by='player_name')

right_handed_batters = all_data |> 
  filter(player_name!="Gallo, Joey") |> 
  filter(launch_angle < 10) |> 
  summarize(N = n(),
            Pulled = 100 * mean(hc_x < 125.42, na.rm = TRUE),
            Shift = 100 * 
              mean(if_fielding_alignment == "Infield shift",
                   na.rm = TRUE),.by='player_name') 

rbind(left_handed_batters, right_handed_batters)
```

What does this tell us about the tendencies of these batters and how defenses approached them?


