---
title: "MA388 Sabermetrics: Lesson 18"
subtitle: "Comparing Peak Trajectories - Part II"
author: "Your Name Here"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Comparing Trajectories of Home Run Hitters

1. Find the ten players in baseball history who have hit the most career home runs.

```{r}
library(tidyverse)
library(Lahman)
library(knitr)
```

```{r}
# ***Your code here.***

```

2. Fit quadratic functions to the home run rates of the ten players, where HR.RATE = HR / AB.  Display the fitted trajectories in a single figure.

```{r}
# ***Your code here.***

```

3. On the basis of your work, which player had the highest estimated home run rate at his peak?  Which player among the ten had the smallest peak home run rate?

```{r}
# ***Your code here.***

```

*Your answer here.*

4. Do any of the players have unusual career trajectory shapes?  Is there any possible explanation for these unusual shapes?

*Your answer here.*


