---
title: "MA388 Sabermetrics: Lesson 35"
subtitle: "Shiny Apps"
author: "LTC Jim Pleuss"
format:
  html:
    theme: cosmo
    toc: true
  pdf:
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE, message= FALSE, warning = FALSE)
```

## Purpose

Last class we focused on generating slides using Quarto, and we explored ways to scale up your production to generate a set of slides for any combination of teams, years, etc.  

What are some benefits of this capability?

- Easily generate products at scale using a proven analysis pipeline.

- You can email your generated product.  There's no web hosting.

What are some limitations?

- If your product doesn't answer the reader's question, the reader has no way to refine the product to answer their question of interest.

- The reader has very limited opportunity to interact with the data.  The data doesn't automatically travel with the static document you've created (possibly good or bad).

## Shiny Apps!

A Shiny app is fundamentally different.  Now you'll host a small (though it could be quite large) application on a web server, and people will visit it via a published URL.  It's pretty simple:

- Allow the user to select some permissible values of different parameters you'll use to control any analysis you present.

- Lay out a user interface (UI) and define server functions that turn user selections into tables, plots, etc.

## Starting Example

**Step 1:** File -> New File -> Shiny Web App

Give the app a name with no spaces.  I called mine "BattingLeaders" and selected the "Single File" option (default) and specified a directory where I wanted the app to live.  I now have a folder in this locaiton called "BattingLeaders" with a single file titled "app.R" inside.

**Step 2:** Press the "Run App" button near the top right of the app editor.  Drag the slider to different values and see what happens.

All the basics are here - a title, an input widget, and a reactive plot that changes as you change the input slider.  This is basically it for Shiny Apps -- you mix static content, user inputs, and reactive content.

**Step 3:** Explore the code that produced this app.

You should see a section called the `ui` that has elements you can connect to what you see in the app - a layout, some panels, a slider input, and a plot output.  

Now look at the `server` section.  There's a single `output` element named `distplot` that is created in the server and made visible in the `ui` section as an argument to the `plotOutput` function.  Notice that the plot code contained in this element would work anywhere with one exception -- it is reliant on knowing what `input$bins` equals.  You can guess where it gets that value -- it's from the `sliderInput` object named "bins" (the first element of an input object definition is its name/ID).

**Step 4:** Tweak this app to do something else.

- Add static text.

After the `titlePanel` line in the `ui`, add this as a new line (including the comma):

`h6("This is some really helpful explanation of what you're looking at."),`

Run the app again to see what happened.

- Add an input.

Put a comma at the end of the `sliderInput` definition after `value=30)`.  On the next line, add this text object:

`h6("I couldn't think of anything good to add here, so I'm switching data sets:"),` (include the comma)

...followed by this new input object:

```{r, eval = FALSE}
radioButtons("cylinders", 
             label = h3("Select a number of cylinders to explore in mtcars:"),
             choices = list("4 cylinders" = 4, 
                            "6 cylinders" = 6, 
                            "8 cylinders" = 8), 
             selected = 4)
```

How do you learn what input widgets are available and how to employ them?  Visit the [Shiny Widget Gallery](https://shiny.posit.co/r/gallery/widgets/widget-gallery/).

Run the app again to see what happened.

- Add an output.

Let's add a plot to the app that does something with `mtcars` and your input selection.  Copy and paste the `output$distPlot` code block immediately underneath the existing code block (no need for a comma between the two plot blocks).  Call this second plot block `output$distPlot2` and add a second `plotOutput` object to the `ui` section like this:

```{r, eval = FALSE}
mainPanel(
  plotOutput("distPlot"),
  plotOutput("distPlot2"),
)
```

Run the app again to see what happened.  We should see our cylinders radio buttons and a second inter-eruption distribution plot.  It's time to connect these objects.  First, add `library(tidyverse)` to the top of the app by the `library(shiny)` call.  Next, change `distPlot2` as follows:

```{r, eval = FALSE}
output$distPlot2 <- renderPlot({
  mtcars |> 
    filter(cyl == input$cylinders) |> 
    ggplot(aes(x = mpg)) + 
    geom_histogram(bins = 6) + 
    labs(x = 'MPG',
         title = paste('Histogram of MPG for vehicles with', 
                       input$cylinders, 
                       "cylinders"))
})
```

**That's it!  This basic knowledge -- plus tons of creativity and trial and error -- is what you need to make a much cooler, much more complicated app.**

## Challenge Problem

Make a Shiny App that allows the user to select a starting year and ending year between 1970 and 2023 as well as a choice of batting stats (start with batting average and home run total).  

Display a histogram of all players' stats during the time period of interest, showing only the histogram of the selected batting stat.
