---
title: "MA388 Sabermetrics: Lesson 21"
subtitle: "Simulating Baseball - Part I"
format:
  html:
    theme: cosmo
    toc: true
  pdf:
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## Markov Chains


Markov chains model the probabilities of moving between different states. I want you to draw a state diagram (think a network diagram with states as circles and lines connecting the states with probabilities) for a baseball half-inning based on the number of outs where probabilities reflect the state transitions after one plate appearance. Assume that every plate appearance has a .7 probability of resulting in an out and .3 probability of not and there is never more than one out for a plate appearance.

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::
\vspace{1in}

*What is the probability of having 0 outs after two plate appearances?*
*What about having two outs?*

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::
\vspace{1in}

What is the memoryless property of Markov chains?  Does this allow for "momentum"?


::: {.html}
<div style="margin-top: 0.5in;"></div>
:::
\vspace{1in}

So this was all very basic, but hopefully a useful intro. What if we wanted to model baseball this way to determine how many at-bats occur in a given baseball half-inning (or entire game)? How might you approach this using Markov Chains and the data we have? How would you change the states?

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

\vspace{1in}

```{r}
library(tidyverse)
library(knitr)
library(baseballr)

# Load the data.
retro2024 <- 
retrosheet_data(years_to_acquire = '2024') |> 
  pluck('2024') |> 
  pluck('events')
```

Let's add states, etc., similarly to how we've done it in previous lessons.  One important difference is that we don't care if/where baserunners were left on base when the third out was made.  The third out is, however, an important state for us to capture.

\newpage

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

```{r, cache = TRUE}
# Add states, etc.  
retro2024 <- retro2024 |> 
  mutate(runs_before = away_score_ct + home_score_ct,
         half_inning = paste(game_id, inn_ct, bat_home_id),
         runs_scored = 
           (bat_dest_id > 3) + (run1_dest_id > 3) + 
           (run2_dest_id > 3) + (run3_dest_id > 3))

half_innings <- retro2024 |>
  summarize(outs_inning = sum(event_outs_ct),
            runs_inning = sum(runs_scored),
            runs_start = first(runs_before),
            max_runs = runs_inning + runs_start, .by ='half_inning')

retro2024_complete <- retro2024 |>
  inner_join(half_innings, by = "half_inning") |>
  mutate(runs_roi = max_runs - runs_before) |>
  mutate(bases = 
           paste(ifelse(base1_run_id > '',1,0),
                 ifelse(base2_run_id > '',1,0),
                 ifelse(base3_run_id > '',1,0), sep = ""),
         state = paste(bases, outs_ct)) |>
  mutate(is_runner1 =
           as.numeric(run1_dest_id==1 | bat_dest_id == 1),
         is_runner2 = 
           as.numeric(run1_dest_id == 2 | run2_dest_id == 2 |
                        bat_dest_id == 2),
         is_runner3 = 
           as.numeric(run1_dest_id == 3 | run2_dest_id == 3 |
                        run3_dest_id == 3 | bat_dest_id == 3),
         new_outs = outs_ct + event_outs_ct,
         new_bases = paste(is_runner1, is_runner2, is_runner3, sep = ""),
         new_state = paste(new_bases, new_outs)) |>
  filter((state != new_state) | (runs_scored > 0)) |>
  filter(outs_inning == 3, bat_event_fl == TRUE) |> 
  mutate(new_state = str_replace(new_state, "[0-1]{3} 3", "3"))
```

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

Let's look at our resulting data frame:

\newpage

```{r}
retro2024_complete |> 
  select(game_id, bat_id, state, new_state) |> 
  head(10) |> 
  kable()
```

We're now ready to compute a matrix of transition probabilities. How many starting states are there? What about ending states? How can we compute these state-to-state probabilities?

\vspace{1in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

```{r}
T_matrix <- retro2024_complete  |> 
  select(state, new_state) |> 
  table()
T_matrix
```

```{r}
P_matrix <- prop.table(T_matrix, 1)
dim(P_matrix)
```

Once you get to three outs, there's nowhere left to go -- you will stay in three outs with probability 1.
```{r}

P_matrix <- P_matrix |>
  rbind("3" = c(rep(0, 24), 1))
# dim(P_matrix)
P_matrix |> round(3)
```

\newpage

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

How can we check the validity of this transition matrix?  Pick a state -- what should the sum of all transition probabilities from that state to any other be?

\vspace{1in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

```{r}
P_matrix |>
  apply(MARGIN = 1, FUN = sum)
```

Let's examine the nobody-on, nobody-out state.  What are the most likely resulting states?

```{r}
P_matrix |> 
  as_tibble(rownames = "state")  |> 
  filter(state == "000 0") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |> 
  filter(Prob > 0) |> 
  kable()
```

What about with a runner on 2nd base and two outs?

```{r}
P_matrix |>
  as_tibble(rownames = "state") |>
  filter(state == "010 2") |>
  pivot_longer(
    cols = -state, 
    names_to = "new_state", 
    values_to = "Prob" 
  ) |>
  filter(Prob > 0) |> 
  kable()
```

So far, so good...  Now we need to actually simulate the Markov Chain.  This involves being in a state and transitioning into possible other states in a probabilistic manner...a large number of times.

The key equation for keeping score is:

$$
runs = \left(N_{runners}^{(before)} + Outs^{(before)} + 1\right) - \left(N_{runners}^{(after)} + Outs^{(after)}\right)
$$

```{r}
# This function essentially "adds" the state: "011 1" becomes 3.
num_havent_scored <- function(s) {
  s |>
    str_split("") |>
    pluck(1) |>
    as.numeric() |>
    sum(na.rm = TRUE)
}

# "Add" the state (runners + outs) for every state.
runners_out <- T_matrix |>
  row.names() |>
  set_names() |>
  map_int(num_havent_scored)

# This operation computes the runs equation above for every state-to-state 
# transition, including the impossible transitions that we can safely ignore.
R_runs <- outer(
  runners_out + 1, 
  runners_out, 
  FUN = "-"
) |> 
  cbind("3" = rep(0, 24))
```

Before we go any further, what are the key tasks for simulating a half-inning of baseball?

\vspace{1in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

Here we go - let's simulate!

```{r}
# This function returns the number of runs scored in a simulated half-inning.
simulate_half_inning <- function(P, R, start = 1) {
  s <- start
  path <- NULL
  runs <- 0
  while (s < 25) {
    s_new <- sample(1:25, size = 1, prob = P[s, ])
    path <- c(path, s_new)
    runs <- runs + R[s, s_new]
    s <- s_new
  }
  runs
}
```

Let's simulate 10,000 half-innings.

```{r}
set.seed(388)
simulated_runs <- 1:10000 |>
  map_int(~simulate_half_inning(P_matrix, R_runs))
P_matrix[1,]
```

What was the observed distribution of half-inning run totals?

```{r}
table(simulated_runs)
```

How often do teams score at least five runs in an inning?

```{r}
sum(simulated_runs >= 5) / 10000
```

What is the average number of runs scored per half-inning?

```{r}
mean(simulated_runs)
```

What have we left out?  Name some non-batting plays that have the potential to change the state and potentially score runs.

\vspace{1in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::
