---
title: "MA388 Sabermetrics: Lesson 28"
subtitle: "Home Run Hitting I"
format:
  html:
    theme: cosmo
    toc: true
  pdf:
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
---


```{r}
library(tidyverse)
library(knitr)
library(baseballr)
```

## Review

**In a logistic model for hit probability, is it appropriate to use a linear function of launch angle and exit velocity? Explain.**

\vspace{.75in}

**In the DiMaggio streak example, explain how and why we used simulation.**

\vspace{.75in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

## What's with all these home runs?

[Video](https://www.mlb.com/news/mlb-report-on-baseballs-home-run-rates-c278120310)

Here is what they are talking about:

```{r, fig.height=3}
library(tidyverse)
library(Lahman)

HRrates <- Teams |> 
  filter(yearID > 1979) |> 
  group_by(yearID) |> 
  summarize(HR = sum(HR),
            AB = sum(AB)) |> 
  mutate(rate = HR/AB)

HRrates |> 
  ggplot(aes(x = yearID, y = rate)) +
  geom_line() + 
  labs(title = "HR Rate (HR/AB) over Time", x = "Year", y = "HR Rate")
```

### Some potential explanations:

1. **Explanation 1.**  The additional home runs are due to changes in the baseball.

2. **Explanation 2.**  The additional home runs are due to changes in the hitters.

Today, we are going to assess which explanation is more compatible with the data.

### Here is the approach:

1. Fit a model for home run probability as a function of launch speed and angle **in the 2015 season**.

2. Use the model to estimate the predicted probability of a home run on each ball put in play **in the 2017 season**.

3. Use the estimated probabilities to simulate 1,000 seasons.

4. Compare the home runs in the simulated seasons to the observed home runs in 2017.

### What does this tell us?

*If the observed rate in 2017 is consistent with the simulated results, which explanation does this support? Why?*

\vspace{1in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

*What other factors should we consider?*

\vspace{1in}

::: {.html}
<div style="margin-top: 0.5in;"></div>
:::

### Fit a model for home run probability in 2015.

```{r}
library(tidyverse)
library(baseballr)
library(lubridate)
```

Obtain all pitches in 2015 in which the ball was put in play.

```{r}
hit_into_play2015 <- read_csv("statcast_2015_bip.csv") |> 
  filter(launch_speed < 150,
         launch_speed > 0,
         launch_angle < 360)
```

Fit a model for home run probability using the 2015 data. 

```{r}
library(mgcv)
fit <- hit_into_play2015 |> 
  # sample_frac(size=.5) |> 
  gam(events == "home_run" ~ s(launch_speed, launch_angle),
      family = binomial,
      data = _)
```

Visualize the predictions.

```{r, fig.height=3}
library(broom)
hit_into_play2015 <- fit |> 
  augment(type.predict = "response",
          newdata = hit_into_play2015) |> 
  rename(HRprob = .fitted)

hr_prob_2015_fig <- 
  hit_into_play2015 |> 
  ggplot(aes(x = launch_speed, y = launch_angle, color = HRprob)) +  
  geom_point(alpha = 0.2) +
  scale_color_gradient("HR prob", low = "lightgray", high = "blue") +
  theme_minimal()+
  labs(x='Launch speed (mph)',y='Launch angle (degrees)')

hr_prob_2015_fig
```

Let's zoom in.

```{r, fig.height=3}
hit_into_play2015 |> 
  filter(launch_speed > 90, launch_speed < 115,
         launch_angle > 15, launch_angle < 45) |> 
  ggplot(aes(x = launch_speed, y = launch_angle, color = HRprob)) +  
  geom_point() +
  scale_color_gradient("HR prob", low = "lightgray", high = "blue")  + 
  theme_minimal()
```

That's not really pretty, let's try a grid.

```{r, fig.height = 3}
grid <- expand.grid(launch_speed = seq(90, 115, length.out = 50),
                    launch_angle = seq(15,45,length.out = 50))

grid_predictions <- fit |> 
  augment(type.predict = "response",
          newdata = grid) |> 
  rename(HRprob = .fitted)

grid_predictions  |>  ggplot(aes(x = launch_speed, y = launch_angle)) +  
  geom_tile(aes(fill = HRprob)) + 
  geom_contour(aes(z = HRprob), breaks = seq(from = 0.1, to = 0.9, by = 0.2)) + 
  xlab("Exit velocity (mph)") + 
  ylab("Launch angle (degrees)") +  
  scale_fill_gradient("HR prob", low = "lightgray", high = "blue")  + 
  theme_minimal()
```

### Use the 2015 model to predict on 2017.

Obtain all pitches in 2017 in which the ball was put in play.

```{r}
hit_into_play2017 <- read_csv("statcast_2017_bip.csv") |> 
  filter(launch_speed < 150,
         launch_speed > 0,
         launch_angle < 360)

hit_into_play2017 <- fit |> 
  augment(type.predict = "response", newdata = hit_into_play2017) |>
  rename(HRprob = .fitted,HRprobse = .se.fit)

hit_into_play2017 |> 
  select(launch_speed, launch_angle, HRprob) |> 
  head(20) |> 
  knitr::kable()
```

### Simulate 1,000 seasons.

```{r}
sim_hr <- function(){
  N <- nrow(hit_into_play2017)
  sum(runif(N) < hit_into_play2017$HRprob, na.rm = TRUE)
}

HR_predict <- tibble(HR = replicate(1000, sim_hr()))

HR_actual <- hit_into_play2017 |> 
  count(events) |> 
  filter(events == "home_run") |> 
  pull(n)

HR_predict |> ggplot(aes(HR)) +  
  geom_histogram(color = "black", fill = "white", bins = 20) +
  geom_vline(xintercept = HR_actual, color = "blue", size = 2) +  
  annotate("text", x=5550, y=175, label = "Actual HR") + 
  labs(x = "Predicted Number of Home Runs",
       y = "Count") 
```

**What do the results of this simulation tell us?**

\vspace{1in}


```{r}
fit2017 <- hit_into_play2017 |>
  gam(events == "home_run" ~ s(launch_speed, launch_angle),
      family = binomial,
      data = _)


hit_into_play2017 <- fit2017 |> 
  augment(type.predict = "response", newdata = hit_into_play2017) |> 
  rename(HRprob2017data = .fitted, HRprob2017se = .se.fit)

hr_prob_change_fig <- 
hit_into_play2017  |>  
  mutate(model_diff = HRprob2017data-HRprob) |> 
  # filter(!between(model_diff,-.05,.05)) |> 
  ggplot(aes(x = launch_speed, y = launch_angle, color=model_diff)) +  
  geom_point(alpha=1)+
    # geom_tile(aes(fill = model_diff)) + 
  # geom_contour(aes(z = model_diff), breaks = seq(from = 0.1, to = 0.9, by = 0.2)) + 
  xlab("Launch speed (mph)") + 
  ylab("Launch angle (degrees)") +  
  scale_color_gradient2("HR Prob Delta", low = "gray",mid='white', high = "blue",midpoint = 0)  + 
  theme_minimal()

library(patchwork)
hr_prob_2015_fig + hr_prob_change_fig
```

