---
title: "MA388 Sabermetrics: Lesson 19"
subtitle: "Comparing Peak Trajectories - Part II"
author: "LTC Jim Pleuss"
format:
  html:
    theme: cosmo
    toc: true
  pdf:
    documentclass: article
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Comparing Trajectories of Home Run Hitters

1. Find the ten players in baseball history who have hit the most career home runs.

```{r}
library(tidyverse)
library(Lahman)
library(knitr)
```

```{r}

Top10 <- Batting |> 
  summarize(totalHR = sum(HR), .by ='playerID') |> 
  slice_max(totalHR,n=10) |> 
  inner_join(People, by = "playerID") |> 
  mutate(label = str_c(nameFirst,nameLast,sep=' ')) |> 
  relocate(label,totalHR)

Top10 |> 
  select(label, totalHR, debut) |> 
  kable()
  
```

2. Fit quadratic functions to the home run rates of the ten players, where HR.RATE = HR / AB.  Display the fitted trajectories in a single figure.

```{r}

top10_stats <- Batting |> 
  filter(playerID %in% Top10$playerID) |> 
  summarize(HR.RATE = sum(HR)/sum(AB),.by=c('playerID','yearID')) |> 
  inner_join(People, by = "playerID") |> 
  mutate(birthyear = ifelse(birthMonth >= 7,
                            birthYear + 1, birthYear),
         label = str_c(nameFirst,nameLast,sep=' '), 
         Age = yearID - birthyear)

top10_stats |> 
  ggplot(aes(x = Age, y = HR.RATE)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y~poly(x,2), se = FALSE) + 
  facet_wrap(~label, ncol = 5)
```

3. On the basis of your work, which player had the highest estimated home run rate at his peak?  Which player among the ten had the smallest peak home run rate?

```{r}

top10_model <- function(this_playerID){
  
  this_player_stats <- top10_stats |> 
    filter(playerID == this_playerID)
  
  fit <- lm(HR.RATE ~ I(Age - 30) + I((Age - 30)^2), data = this_player_stats)
  
  A <- coef(fit)[1]
  B <- coef(fit)[2]
  C <- coef(fit)[3]
  
  # Return all the stuff you just obtained.
  tibble(label = head(this_player_stats$label,1), playerID = this_playerID, A, B, C,
         age.max = 30 - B / C / 2, 
         max = A - B^2 / C / 4)
}

# Map the model function to the list of playerIDs you obtained in 1.
HR_results <- map_df(Top10$playerID,top10_model)
HR_results |> 
  kable()
```

```{r}
# Highest estimated max HR rate:
HR_results |> 
  slice_max(max)
```

```{r}
# Lowest estimated max HR rate:
HR_results |> 
  slice_min(max) 
```

*Your answer here.*

4. Do any of the players have unusual career trajectory shapes?  Is there any possible explanation for these unusual shapes?

*Your answer here.*

